---
title: "Spatial Transcriptomics Non-stationarity"
author: "Puneet Velidi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 6
knitr:
  opts_chunk:
    cache: true
---

```{r setup, include=FALSE}
library(SpatialExperiment)
library(ExperimentHub)
library(ggplot2)
library(dplyr)
library(tidyr)
library(Matrix)

# Load analysis functions (INLA-based spatial models and Bayes factor)
source("hpc/analysis_functions.R")

eh <- ExperimentHub()
```

# Statistical Model

## Model Specification

For each spatially variable gene we fit two SPDE-based spatial models (Lindgren et al. 2011):

$$Y(s) = \mu + u(s) + \epsilon(s)$$

where $u(s)$ is a Gaussian Markov random field with Matérn covariance defined
through the SPDE

$$(\kappa^2(s) - \Delta)^{\alpha/2}(\tau(s)\, u(s)) = W(s), \quad \alpha = 2,$$

with $\nu = 1$ smoothness and $\epsilon(s) \sim N(0, \sigma_\epsilon^2)$ iid
nugget.  The marginal variance and practical range of $u$ are:

$$\sigma_b^2 = \frac{1}{4\pi\,\kappa^2\tau^2}, \qquad
  \rho = \frac{\sqrt{8}}{\kappa}.$$

**Stationary model** $M_0$: $\log\kappa(s) = \theta_2$ and $\log\tau(s) = \theta_1$
are spatially constant.

**Non-stationary model** $M_1$ (Ingebrigtsen et al. 2014):
$\log\tau(s) = \theta_1$ (constant) but the range parameter varies linearly
across the tissue:

$$\log\kappa(s) = \theta_2 + \theta_3\,\tilde x(s) + \theta_4\,\tilde y(s),$$

where $\tilde x, \tilde y$ are standardised spatial coordinates.
This is implemented via spatially-varying $\mathbf{B}_\kappa$ matrices in
`inla.spde2.matern`.

Both models use `inla.spde2.matern` with identical prior specifications so the
marginal log-likelihoods are directly comparable.

## Bayes Factor

INLA provides the marginal log-likelihood $\log p(Y \mid M)$ via the
integrated nested Laplace approximation.  The Bayes factor comparing $M_1$
to $M_0$ is:

$$\log BF_{10} = \log p(Y \mid M_1) - \log p(Y \mid M_0).$$

| $\log BF_{10}$ | Evidence |
|---|---|
| $(-\log 3,\; \log 3)$ | Anecdotal |
| $(\log 3,\; \log 10)$ | Moderate for $M_1$ |
| $(\log 10,\; \log 30)$ | Strong for $M_1$ |
| $(\log 30,\; \log 100)$ | Very strong for $M_1$ |
| $> \log 100$ | Extreme for $M_1$ |

Negative values give symmetric evidence for $M_0$ (stationary).

## Proportion of Spatial Variability

$$\text{PSV} = \frac{\sigma_b^2}{\sigma_b^2 + \sigma_\epsilon^2}$$

---

# Example Analysis: Human Breast Cancer IDC

```{r load-data, cache=TRUE}
visium_hbc <- query(eh, "HumanBreastCancerIDC")
spe <- visium_hbc[[1]]
stopifnot(inherits(spe, "SpatialExperiment"))

# Ensure gene_name column exists (required by nnSVG)
rd <- rowData(spe)
if (!"gene_name" %in% colnames(rd)) {
  col <- intersect(c("symbol", "Symbol", "SYMBOL"), colnames(rd))[1]
  rowData(spe)$gene_name <- if (!is.na(col)) rd[[col]] else rownames(spe)
}
```

```{r filter-normalize, cache=TRUE}
counts_mat <- counts(spe)
n_spots    <- ncol(spe)

# Keep genes detected in >= 0.5% of spots with >= 3 total counts
keep <- (Matrix::rowSums(counts_mat > 0) >= ceiling(n_spots * 0.005)) &
        (Matrix::rowSums(counts_mat) >= 3)
spe_filtered <- spe[keep, ]
cat("Retained", sum(keep), "/", length(keep), "genes\n")

# Library-size normalisation and log2 transform
lib_sizes  <- Matrix::colSums(counts(spe_filtered))
sf         <- lib_sizes / mean(lib_sizes)
assay(spe_filtered, "logcounts") <- log2(t(t(counts(spe_filtered)) / sf) + 1)
```

```{r run-bf-analysis, cache=TRUE}
gene_indices <- seq_len(nrow(spe_filtered))
rd2          <- rowData(spe_filtered)
gene_names   <- if ("gene_name" %in% colnames(rd2)) as.character(rd2$gene_name)
                else if ("symbol" %in% colnames(rd2)) as.character(rd2$symbol)
                else rownames(spe_filtered)

coords <- as.data.frame(spatialCoords(spe_filtered))
colnames(coords)[1:2] <- c("x", "y")

cat("Analysing all", length(gene_indices), "expressed genes...\n")
cat("(For parallel execution across all datasets use hpc/run_analysis.R)\n\n")

results_list <- vector("list", length(gene_indices))

for (i in seq_along(gene_indices)) {
  cat(sprintf("[%d/%d] %s\n", i, length(gene_indices), gene_names[i]))
  expr <- log1p(counts(spe_filtered)[gene_indices[i], ])
  results_list[[i]] <- tryCatch(
    analyze_gene_nonstationarity(
      coords         = coords,
      expr           = expr,
      gene_name      = gene_names[i],
      normalize      = TRUE,
      detrend        = TRUE,
      detrend_method = "polynomial",
      verbose        = FALSE
    ),
    error = function(e) {
      message("Error for gene ", gene_names[i], ": ", e$message)
      NULL
    }
  )
}

results_df <- do.call(rbind, Filter(Negate(is.null), results_list))
rownames(results_df) <- NULL
cat("\nCompleted:", nrow(results_df), "genes\n")
```

---

# Results

```{r show-bf-table}
print(results_df[, c("gene_name", "log_bayes_factor", "bayes_factor",
                      "bf_interpretation", "prop_spatial_stationary")])
```

```{r results-summary}
summary_stats <- results_df %>%
  filter(!is.na(log_bayes_factor)) %>%
  summarise(
    n_genes                    = n(),
    prop_favor_nonstationary   = mean(log_bayes_factor > 0),
    mean_log_bf                = mean(log_bayes_factor),
    median_log_bf              = median(log_bayes_factor),
    mean_psv_stationary        = mean(prop_spatial_stationary, na.rm = TRUE),
    mean_psv_nonstationary     = mean(prop_spatial_nonstationary, na.rm = TRUE)
  )
print(summary_stats)
```

```{r plot-bf-distribution}
ggplot(results_df, aes(x = log_bayes_factor)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  geom_vline(xintercept =  log(10), linetype = "dotted", colour = "orange") +
  geom_vline(xintercept = -log(10), linetype = "dotted", colour = "orange") +
  labs(
    title    = "Distribution of Log Bayes Factors",
    subtitle = paste0("Positive = non-stationary covariance kernel favoured",
                      "  |  dashed = BF=1, dotted = BF=10"),
    x = expression(log~BF[10]),
    y = "Number of genes"
  ) +
  theme_minimal()
```

```{r plot-psv-comparison}
comparison_df <- results_df %>%
  select(gene_name, prop_spatial_stationary, prop_spatial_nonstationary) %>%
  pivot_longer(
    cols      = c(prop_spatial_stationary, prop_spatial_nonstationary),
    names_to  = "model",
    values_to = "psv"
  ) %>%
  mutate(model = recode(model,
    prop_spatial_stationary    = "Stationary",
    prop_spatial_nonstationary = "Non-stationary"
  ))

ggplot(comparison_df, aes(x = model, y = psv, fill = model)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Stationary" = "steelblue",
                               "Non-stationary" = "coral")) +
  labs(
    title    = "Proportion of Spatial Variability by Model",
    x        = NULL,
    y        = expression(sigma[b]^2 / (sigma[b]^2 + sigma[epsilon]^2))
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Notes

1. **Normalisation**: Quantile normalisation to standard normal before fitting.

2. **Detrending**: Second-order polynomial detrending removes large-scale mean
   trends so the SPDE captures residual spatial covariance structure.

3. **Stationary model** ($M_0$): Two hyperparameters—$\log\tau$ and
   $\log\kappa$—are spatially constant.

4. **Non-stationary model** ($M_1$): Four hyperparameters.  The range
   $\rho(s) = \sqrt{8}/\kappa(s)$ varies linearly across the tissue through
   the B-matrix formulation of Ingebrigtsen et al. (2014).

5. **Bayes factor validity**: Both models use `inla.spde2.matern` with the
   same prior specification, so the marginal log-likelihoods are directly
   comparable and the BF is a principled model-selection criterion.

6. **HPC runs**: For full-scale analysis across all datasets and all spatially
   variable genes, use `hpc/run_analysis.R` submitted via
   `hpc/submit_analysis.sbatch`.

## References

- Lindgren, F., Rue, H. & Lindström, J. (2011). An explicit link between
  Gaussian fields and Gaussian Markov random fields: the SPDE approach.
  *JRSS-B*, 73(4), 423–498.

- Ingebrigtsen, R., Lindgren, F. & Steinsland, I. (2014). Spatial models
  with explanatory variables in the dependence structure.
  *Spat. Stat.*, 8, 20–38.

- Kass, R.E. & Raftery, A.E. (1995). Bayes factors.
  *JASA*, 90(430), 773–795.
