---
title: "Spatial Transcriptomics Non-stationarity"
author: "Puneet Velidi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 6
knitr:
  opts_chunk: 
    cache: true
---

```{r setup, include=FALSE}
library(SpatialExperiment)
library(ExperimentHub)
library(ggplot2)
library(gstat)
library(dplyr)
library(tidyr)

# Check INLA installation and load
# Note: INLA requires restarting R after installation for binaries to load
inla_available <- FALSE
inla_error_msg <- NULL

if (require("INLA", quietly = TRUE)) {
  # Test INLA functionality - this will catch binary loading issues
  tryCatch({
    # Simple test to see if INLA binaries are loaded
    test_result <- INLA::inla.mesh.1d(seq(0, 1, length.out = 5))
    inla_available <- TRUE
    cat("✓ INLA is properly installed and binaries are loaded.\n")
  }, error = function(e) {
    inla_error_msg <<- e$message
    warning("⚠ INLA is installed but binaries are not loading properly.\n",
            "Error: ", e$message, "\n",
            "SOLUTION: Please RESTART R/RStudio completely and try again.\n",
            "INLA requires a restart after installation for binaries to load.\n",
            "The analysis functions will not work until INLA binaries are available.")
  })
} else {
  warning("⚠ INLA package is not installed.\n",
          "Please install it using:\n",
          "install.packages('INLA', repos=c(getOption('repos'), INLA='https://inla.r-inla-download.org/R/stable'), dep=TRUE)\n",
          "Then RESTART R/RStudio and try again.")
}

# Store INLA availability status for use in analysis functions
assign(".inla_available", inla_available, envir = .GlobalEnv)

eh <- ExperimentHub()
```

```{r load-datasets, cache=TRUE}
# Load datasets for analysis
# Uncomment below to load multiple Visium datasets:
#visium_all <- query(eh, c("SpatialExperiment", "Visium"))
#visium_datasets <- lapply(seq_along(visium_all), function(i) visium_all[[i]])

# Load Human Breast Cancer dataset as example
# Note: This chunk queries for the dataset but doesn't extract it yet
# The example-analysis chunk below shows how to properly load and use it
visium_hbc <- query(eh, "HumanBreastCancerIDC")

# Example: extract coordinates for a specific gene (if needed for visualization)
# Uncomment below to use:
# if (length(visium_hbc) > 0) {
#   spe <- visium_hbc[[1]]
#   if (inherits(spe, "SpatialExperiment")) {
#     gene_idx <- 100
#     coords <- as.data.frame(spatialCoords(spe))
#     coords$expr <- log1p(counts(spe)[gene_idx, ])
#     colnames(coords)[1:2] <- c("x", "y")
#   }
# }
```

```{r}
# This chunk is for quick visualization/testing with a single gene
# For the main analysis, use the analyze_multiple_genes() function below
# Uncomment below to use:
# if (exists("visium_hbc") && length(visium_hbc) > 0) {
#   spe <- visium_hbc[[1]]
#   if (inherits(spe, "SpatialExperiment")) {
#     gene_idx <- 100
#     coords <- as.data.frame(spatialCoords(spe))
#     coords$expr <- log1p(counts(spe)[gene_idx, ])  # Select specific gene
#     colnames(coords)[1:2] <- c("x", "y")
#   }
# }
```

```{r visualize, eval=FALSE}
# Visualization chunk - uncomment and modify if you want to visualize specific genes
# This requires visium_datasets to be loaded (currently commented out above)
# 
# for (i in seq_along(visium_datasets)) {
#   spe <- visium_datasets[[i]]
#   gene_idx <- 100
#   rd <- rowData(spe)
#   gene_name <- if ("symbol" %in% colnames(rd)) rd$symbol[gene_idx] else
#                if ("gene_name" %in% colnames(rd)) rd$gene_name[gene_idx] else
#                gsub("ENS.*G[0-9]+|.*_", "", rownames(spe)[gene_idx])
#   cat("Dataset", i, "- Gene:", gene_name, "\n")
# 
#   coords <- as.data.frame(spatialCoords(spe))
#   coords$expr <- log1p(counts(spe)[gene_idx, ])
#   colnames(coords)[1:2] <- c("x", "y")
# 
#   eh_meta <- mcols(visium_all)[i, ]
#   title <- paste0(eh_meta$title, "\nSpecies: ", ifelse(is.na(eh_meta$species), "Unknown", eh_meta$species), " | Gene: ", gene_name)
# 
#   print(ggplot(coords, aes(x, y, color = expr)) +
#     geom_point(size = 1, alpha = 0.9) +
#     scale_color_viridis_c(name = "log(expr+1)") +
#     labs(title = title) +
#     theme_minimal() +
#     coord_fixed())
# }

# Example visualization for current dataset (uncomment to use):
# if (exists("spe")) {
#   gene_idx <- 100
#   rd <- rowData(spe)
#   gene_name <- if ("symbol" %in% colnames(rd)) rd$symbol[gene_idx] else
#                if ("gene_name" %in% colnames(rd)) rd$gene_name[gene_idx] else
#                gsub("ENS.*G[0-9]+|.*_", "", rownames(spe)[gene_idx])
#   
#   coords <- as.data.frame(spatialCoords(spe))
#   coords$expr <- log1p(counts(spe)[gene_idx, ])
#   colnames(coords)[1:2] <- c("x", "y")
#   
#   print(ggplot(coords, aes(x, y, color = expr)) +
#     geom_point(size = 1, alpha = 0.9) +
#     scale_color_viridis_c(name = "log(expr+1)") +
#     labs(title = paste("Gene:", gene_name)) +
#     theme_minimal() +
#     coord_fixed())
# }
```

```{r covariograms, eval=FALSE}
# Covariogram visualization chunk - uncomment and modify if needed
# This requires visium_datasets to be loaded (currently commented out above)
#
# for (i in seq_along(visium_datasets)) {
#   spe <- visium_datasets[[i]]
#   gene_idx <- 100
#   rd <- rowData(spe)
#   gene_name <- if ("symbol" %in% colnames(rd)) rd$symbol[gene_idx] else
#                if ("gene_name" %in% colnames(rd)) rd$gene_name[gene_idx] else
#                gsub("ENS.*G[0-9]+|.*_", "", rownames(spe)[gene_idx])
# 
#   coords <- as.data.frame(spatialCoords(spe))
#   coords$expr <- log1p(counts(spe)[gene_idx, ])
#   colnames(coords)[1:2] <- c("x", "y")
# 
#   # Compute covariogram using gstat
#   v <- variogram(expr ~ 1, ~ x + y, coords)
#   v$covariance <- var(coords$expr) - v$gamma
# 
#   eh_meta <- mcols(visium_all)[i, ]
#   print(ggplot(v, aes(dist, covariance)) +
#     geom_point() + geom_line() +
#     labs(title = paste0(eh_meta$title, " | Gene: ", gene_name), x = "Distance", y = "Covariance") +
#     theme_minimal())
# }

# Example covariogram for current dataset (uncomment to use):
# if (exists("spe")) {
#   gene_idx <- 100
#   rd <- rowData(spe)
#   gene_name <- if ("symbol" %in% colnames(rd)) rd$symbol[gene_idx] else
#                if ("gene_name" %in% colnames(rd)) rd$gene_name[gene_idx] else
#                gsub("ENS.*G[0-9]+|.*_", "", rownames(spe)[gene_idx])
#   
#   coords <- as.data.frame(spatialCoords(spe))
#   coords$expr <- log1p(counts(spe)[gene_idx, ])
#   colnames(coords)[1:2] <- c("x", "y")
#   
#   # Compute covariogram using gstat
#   v <- variogram(expr ~ 1, ~ x + y, coords)
#   v$covariance <- var(coords$expr) - v$gamma
#   
#   print(ggplot(v, aes(dist, covariance)) +
#     geom_point() + geom_line() +
#     labs(title = paste("Gene:", gene_name), x = "Distance", y = "Covariance") +
#     theme_minimal())
# }
```

# Nonstationarity Analysis

## ⚠️ IMPORTANT: INLA Installation on macOS Apple Silicon

**The error "INLA installation error; no such file" is a known issue on macOS with Apple Silicon (M1/M2/M3).**

## INLA Installation Troubleshooting

### Solution 1: Install the TESTING version of INLA (Recommended for Apple Silicon)

The stable version may not work on Apple Silicon. Use the **testing** version instead:

```{r install-inla-testing, eval=FALSE}
# Remove existing INLA installation
remove.packages("INLA")

# Install the TESTING version (has better Apple Silicon support)
install.packages("INLA", 
                 repos = c(getOption("repos"), 
                          INLA = "https://inla.r-inla-download.org/R/testing"), 
                 dep = TRUE)

# RESTART R/RStudio after installation
```

### Solution 2: Create symbolic link for INLA binary (if Solution 1 doesn't work)

Run these commands in your Terminal (not R):

``` bash
# Find your INLA installation path - adjust R version as needed
cd ~/Library/R/arm64/4.4/library/INLA/bin/mac

# Create symbolic link from arm64 binary to expected location
mkdir -p 64bit
ln -sf ../mac.arm64/inla 64bit/inla
```

Note: Adjust the path if your R library is in a different location. You can find it by running `.libPaths()` in R.

### Solution 3: Verify installation

After trying the above, restart R and test:

```{r check-inla, eval=FALSE}
# Check if INLA package is found
packageVersion("INLA")

# Try to load INLA
library(INLA)

# Test basic functionality
test_coords <- matrix(c(0, 0, 1, 1, 0.5, 0.5), ncol = 2, byrow = TRUE)
test_mesh <- INLA::inla.mesh.2d(loc = test_coords, max.edge = c(0.5, 1))
cat("INLA is working!\n")
```

### Solution 4: Check INLA binary location

If still having issues, check where INLA expects the binary:

```{r debug-inla, eval=FALSE}
# Find INLA package location
system.file(package = "INLA")

# List files in INLA bin directory
list.files(file.path(system.file(package = "INLA"), "bin", "mac"), recursive = TRUE)
```

## Helper Functions

### Normalization and Detrending

```{r normalization-detrending-functions}
# Function to normalize gene expression data
normalize_expression <- function(expr, method = "quantile") {
  if (method == "quantile") {
    # Quantile normalization
    expr_rank <- rank(expr, ties.method = "average")
    expr_norm <- qnorm((expr_rank - 0.5) / length(expr))
    return(expr_norm)
  } else if (method == "zscore") {
    # Z-score normalization
    return((expr - mean(expr)) / sd(expr))
  } else {
    # Log-normalization (already done typically)
    return(expr)
  }
}

# Function to detrend gene expression data to remove large-scale spatial variability
detrend_expression <- function(coords, expr, method = "polynomial") {
  if (method == "loess") {
    # Use loess smoothing to estimate and remove trend
    loess_fit <- loess(expr ~ x + y, data = coords, span = 0.5)
    trend <- predict(loess_fit, newdata = coords)
    detrended <- expr - trend
    return(list(detrended = detrended, trend = trend))
  } else if (method == "polynomial") {
    # Polynomial detrending
    x_centered <- coords$x - mean(coords$x)
    y_centered <- coords$y - mean(coords$y)
    poly_fit <- lm(expr ~ x_centered + y_centered + 
                   I(x_centered^2) + I(y_centered^2) + I(x_centered * y_centered),
                   data = data.frame(expr = expr, x_centered = x_centered, y_centered = y_centered))
    trend <- predict(poly_fit)
    detrended <- expr - trend
    return(list(detrended = detrended, trend = trend))
  } else {
    # No detrending
    return(list(detrended = expr, trend = rep(0, length(expr))))
  }
}
```

### INLA Spatial Modeling Functions

```{r inla-modeling-functions}
# Function to create mesh for INLA SPDE
create_spde_mesh <- function(coords, max.edge = NULL, cutoff = NULL) {
  # Determine appropriate mesh parameters if not provided
  if (is.null(max.edge)) {
    # Use 1/3 of the range as max.edge
    x_range <- diff(range(coords$x))
    y_range <- diff(range(coords$y))
    max_range <- max(x_range, y_range)
    max.edge <- c(max_range / 5, max_range / 3)
  }
  
  if (is.null(cutoff)) {
    # Set cutoff to avoid too many small edges
    x_range <- diff(range(coords$x))
    y_range <- diff(range(coords$y))
    min_range <- min(x_range, y_range)
    cutoff <- min_range / 20
  }
  
  # Create mesh
  mesh <- inla.mesh.2d(
    loc = as.matrix(coords[, c("x", "y")]),
    max.edge = max.edge,
    cutoff = cutoff
  )
  
  return(mesh)
}

# Function to fit stationary Matern model
# Uses PC priors (inla.spde2.pcmatern) for consistent hyperparameter extraction
fit_stationary_matern <- function(coords, expr, mesh = NULL, verbose = FALSE) {
  if (is.null(mesh)) {
    mesh <- create_spde_mesh(coords)
  }
  
  # Create SPDE object with stationary Matern covariance using PC priors
  # This gives us "Range for spatial" and "Stdev for spatial" hyperparameters
  # which are easier to interpret than the theta parameterization
  spde <- inla.spde2.pcmatern(
    mesh = mesh,
    alpha = 2,  # Smoothness parameter (alpha = nu + d/2, where nu=1 for Matern, d=2 for 2D)
    prior.range = c(0.5, 0.5),  # P(range < 0.5) = 0.5 (weakly informative)
    prior.sigma = c(1, 0.1)     # P(sigma > 1) = 0.1 (weakly informative)
  )
  
  # Create A matrix for projection
  A <- inla.spde.make.A(mesh = mesh, loc = as.matrix(coords[, c("x", "y")]))
  
  # Create index for spatial random effect
  s.index <- inla.spde.make.index(name = "spatial", n.spde = spde$n.spde)
  
  # Create data stack
  stack <- inla.stack(
    data = list(y = expr),
    A = list(A, 1),
    effects = list(
      s.index,
      list(intercept = rep(1, length(expr)))
    ),
    tag = "est"
  )
  
  # Formula: Y(s) = μ + b(s) + ε(s)
  # This is the STATIONARY model - no spatial covariates, just intercept + spatial field
  formula <- y ~ -1 + intercept + f(spatial, model = spde)
  
  # Fit model
  result <- inla(
    formula = formula,
    data = inla.stack.data(stack, spde = spde),
    family = "gaussian",
    control.predictor = list(A = inla.stack.A(stack), compute = TRUE),
    control.compute = list(config = TRUE, dic = TRUE, cpo = TRUE),
    verbose = verbose
  )
  
  return(list(result = result, mesh = mesh, spde = spde, stack = stack))
}

# Function to fit nonstationary Matern model
fit_nonstationary_matern <- function(coords, expr, mesh = NULL, verbose = FALSE) {
  if (is.null(mesh)) {
    mesh <- create_spde_mesh(coords)
  }
  
  # Create SPDE object with nonstationary Matern covariance
  # For nonstationary model, we use inla.spde2.pcmatern with spatially varying parameters
  # This requires specifying prior distributions for the range and variance parameters
  
  # Create nonstationary SPDE
  # Using PC priors: P(range < range0) = alpha_range, P(sigma > sigma0) = alpha_sigma
  spde <- inla.spde2.pcmatern(
    mesh = mesh,
    alpha = 2,  # Smoothness parameter (2 for Matern with smoothness 1)
    prior.range = c(0.5, 0.5),  # P(range < 0.5) = 0.5
    prior.sigma = c(1, 0.1)     # P(sigma > 1) = 0.1
  )
  
  # For nonstationary model, we can make the range parameter spatially varying
  # by using a log-Gaussian process for the range parameter
  # This is more complex and requires additional setup
  
  # Alternative approach: Use a spatially varying variance model
  # Create A matrix
  A <- inla.spde.make.A(mesh = mesh, loc = as.matrix(coords[, c("x", "y")]))
  
  # Create index for spatial random effect
  s.index <- inla.spde.make.index(name = "spatial", n.spde = spde$n.spde)
  
  # For nonstationary model, we can add spatial covariates or use a more flexible SPDE
  # Here we use a model where the variance varies spatially
  # One approach is to use a second spatial field for the log-variance
  
  # Create a second SPDE for the variance field
  spde_var <- inla.spde2.pcmatern(
    mesh = mesh,
    alpha = 2,
    prior.range = c(0.5, 0.5),
    prior.sigma = c(0.5, 0.1)
  )
  
  s.index_var <- inla.spde.make.index(name = "spatial_var", n.spde = spde_var$n.spde)
  
  # Create updated stack with variance field
  stack_ns <- inla.stack(
    data = list(y = expr),
    A = list(A, A, 1),
    effects = list(
      s.index,
      s.index_var,
      list(intercept = rep(1, length(expr)))
    ),
    tag = "est"
  )
  
  # Formula with spatially varying variance
  # This is a simplified nonstationary model
  # A full nonstationary model would require custom INLA code
  formula <- y ~ -1 + intercept + 
    f(spatial, model = spde) +
    f(spatial_var, model = spde_var)
  
  # Fit model
  result <- inla(
    formula = formula,
    data = inla.stack.data(stack_ns, spde = spde, spde_var = spde_var),
    family = "gaussian",
    control.predictor = list(A = inla.stack.A(stack_ns), compute = TRUE),
    control.compute = list(config = TRUE, dic = TRUE, cpo = TRUE),
    verbose = verbose
  )
  
  return(list(result = result, mesh = mesh, spde = spde, spde_var = spde_var, stack = stack_ns))
}

# Alternative simpler nonstationary model using spatially varying range
# This uses a more standard approach with spatially varying hyperparameters
fit_nonstationary_matern_simple <- function(coords, expr, mesh = NULL, verbose = FALSE) {
  if (is.null(mesh)) {
    mesh <- create_spde_mesh(coords)
  }
  
  # Create SPDE with spatially varying parameters
  # We'll use a model where the range parameter varies with spatial location
  # This is achieved by creating a covariate based on spatial coordinates
  
  # Create spatial covariates for nonstationarity
  coords_scaled <- coords
  coords_scaled$x_scaled <- scale(coords$x)[, 1]
  coords_scaled$y_scaled <- scale(coords$y)[, 1]
  
  # Create SPDE
  spde <- inla.spde2.pcmatern(
    mesh = mesh,
    alpha = 2,
    prior.range = c(0.5, 0.5),
    prior.sigma = c(1, 0.1)
  )
  
  # Create A matrix
  A <- inla.spde.make.A(mesh = mesh, loc = as.matrix(coords[, c("x", "y")]))
  
  # Create index for spatial random effect
  s.index <- inla.spde.make.index(name = "spatial", n.spde = spde$n.spde)
  
  # Create data stack with spatial covariates
  stack <- inla.stack(
    data = list(y = expr),
    A = list(A, 1),
    effects = list(
      s.index,
      list(
        intercept = rep(1, length(expr)),
        x_scaled = coords_scaled$x_scaled,
        y_scaled = coords_scaled$y_scaled,
        x2 = coords_scaled$x_scaled^2,
        y2 = coords_scaled$y_scaled^2,
        xy = coords_scaled$x_scaled * coords_scaled$y_scaled
      )
    ),
    tag = "est"
  )
  
  # Formula with spatial covariates that can capture nonstationarity
  # The interaction between spatial field and covariates creates nonstationarity
  formula <- y ~ -1 + intercept + 
    x_scaled + y_scaled + x2 + y2 + xy +
    f(spatial, model = spde)
  
  # Fit model
  result <- inla(
    formula = formula,
    data = inla.stack.data(stack, spde = spde),
    family = "gaussian",
    control.predictor = list(A = inla.stack.A(stack), compute = TRUE),
    control.compute = list(config = TRUE, dic = TRUE, cpo = TRUE),
    verbose = verbose
  )
  
  return(list(result = result, mesh = mesh, spde = spde, stack = stack))
}
```

### Results Extraction Functions

```{r results-extraction-functions}
# Function to extract variance components and calculate spatial variability proportion
extract_variance_components <- function(fit_result) {
  # Extract hyperparameters
  hyper <- fit_result$summary.hyperpar
  
  # For SPDE models with PC priors, the hyperparameters are:
  # - "Range for spatial" (not precision)
  # - "Stdev for spatial" (standard deviation, not precision)
  # - "Precision for the Gaussian observations" (noise)
  
  # Initialize
  sigma_b_sq <- NA
  sigma_eps_sq <- NA
  spatial_range <- NA
  
  # Try to find spatial Stdev (from PC prior SPDE models)
  stdev_idx <- grep("Stdev for spatial", rownames(hyper), ignore.case = TRUE)
  if (length(stdev_idx) > 0) {
    spatial_stdev <- hyper$mean[stdev_idx[1]]
    sigma_b_sq <- spatial_stdev^2  # Variance = Stdev^2
  } else {
    # Try alternative: Precision-based models
    spatial_prec_idx <- grep("Precision for spatial", rownames(hyper))
    if (length(spatial_prec_idx) > 0) {
      spatial_prec <- hyper$mean[spatial_prec_idx]
      sigma_b_sq <- 1 / spatial_prec
    } else {
      # Try any precision with "spatial" in name
      spatial_prec_idx <- grep("Precision.*spatial", rownames(hyper), ignore.case = TRUE)
      if (length(spatial_prec_idx) > 0) {
        spatial_prec <- hyper$mean[spatial_prec_idx[1]]
        sigma_b_sq <- 1 / spatial_prec
      }
    }
  }
  
  # Extract range parameter (useful for interpretation)
  range_idx <- grep("Range for spatial", rownames(hyper), ignore.case = TRUE)
  if (length(range_idx) > 0) {
    spatial_range <- hyper$mean[range_idx[1]]
  }
  
  # Find noise variance (Precision for the Gaussian observations)
  noise_prec_idx <- grep("Precision for the Gaussian observations", rownames(hyper))
  if (length(noise_prec_idx) > 0) {
    noise_prec <- hyper$mean[noise_prec_idx]
    sigma_eps_sq <- 1 / noise_prec
  } else {
    # Try alternative names
    noise_prec_idx <- grep("Precision.*Gaussian|Precision.*observations", rownames(hyper), ignore.case = TRUE)
    if (length(noise_prec_idx) > 0) {
      noise_prec <- hyper$mean[noise_prec_idx[1]]
      sigma_eps_sq <- 1 / noise_prec
    } else {
      sigma_eps_sq <- NA
    }
  }
  
  # Calculate proportion of spatial variability
  if (!is.na(sigma_b_sq) && !is.na(sigma_eps_sq)) {
    prop_spatial <- sigma_b_sq / (sigma_b_sq + sigma_eps_sq)
  } else {
    prop_spatial <- NA
  }
  
  # Extract marginal likelihood (log)
  log_ml <- fit_result$mlik[1, 1]  # Integrated log-likelihood
  
  return(list(
    sigma_b_sq = sigma_b_sq,
    sigma_eps_sq = sigma_eps_sq,
    spatial_range = spatial_range,
    prop_spatial = prop_spatial,
    log_marginal_likelihood = log_ml,
    hyperparameters = hyper
  ))
}

```

## Statistical Models and Bayes Factor

### Model Specification

For each gene, we fit two spatial models of the form:

$$Y(s) = \mu + b(s) + \epsilon(s)$$

where:

-   $Y(s)$ is the (normalized, detrended) gene expression at spatial location $s$
-   $\mu$ is the overall mean (intercept)
-   $b(s)$ is a spatial random effect modeled as a Gaussian process with Matérn covariance
-   $\epsilon(s) \sim N(0, \sigma_\epsilon^2)$ is iid noise (nugget effect)

The spatial effect $b(s)$ is specified as a Whittle-Matérn field via the SPDE approach:

$$b(s) \sim \text{GaussianProcess}(0, C_\theta(s, s'))$$

where $C_\theta$ is the Matérn covariance function with parameters $\theta = (\rho, \sigma_b)$:

-   $\rho$ = spatial range (correlation length scale)
-   $\sigma_b^2$ = marginal variance of the spatial field

**Stationary Model (**$M_0$): The covariance parameters $(\rho, \sigma_b)$ are constant across space.

**Nonstationary Model (**$M_1$): The model includes spatial polynomial covariates $(x, y, x^2, y^2, xy)$ that allow the mean structure to vary spatially, capturing large-scale nonstationarity.

### Variance Components

The **proportion of spatial variability** quantifies how much of the total variance is due to spatial structure:

$$\text{Proportion Spatial} = \frac{\sigma_b^2}{\sigma_b^2 + \sigma_\epsilon^2}$$

where:

-   $\sigma_b^2$ = variance of the spatial effect (from the Matérn field)
-   $\sigma_\epsilon^2$ = variance of the noise/nugget effect

### Bayes Factor Calculation

INLA provides the **marginal likelihood** $p(Y | M)$ for each model via integrated nested Laplace approximation. The Bayes factor comparing the nonstationary model ($M_1$) to the stationary model ($M_0$) is:

$$BF_{10} = \frac{p(Y | M_1)}{p(Y | M_0)}$$

Taking logarithms:

$$\log BF_{10} = \log p(Y | M_1) - \log p(Y | M_0)$$

**Interpretation:**

-   $\log BF_{10} > 0$ (equivalently $BF_{10} > 1$): Evidence favors nonstationary model
-   $\log BF_{10} < 0$ (equivalently $BF_{10} < 1$): Evidence favors stationary model
-   $|\log BF_{10}| > 1$: Substantial evidence
-   $|\log BF_{10}| > 2.3$ (i.e., $BF > 10$): Strong evidence
-   $|\log BF_{10}| > 4.6$ (i.e., $BF > 100$): Decisive evidence

### Bayes Factor Function

```{r bayes-factor-function}
# Function to calculate Bayes factor from marginal likelihoods
# 
# The Bayes factor is:
#   BF_10 = p(Y | M_nonstationary) / p(Y | M_stationary)
#         = exp(log_ml_nonstationary - log_ml_stationary)
#
# log(BF_10) = log p(Y | M_1) - log p(Y | M_0)
#
# Interpretation:
#   BF > 1 (log BF > 0): Favors nonstationary model
#   BF < 1 (log BF < 0): Favors stationary model

calculate_bayes_factor <- function(log_ml_stationary, log_ml_nonstationary) {
  # Log Bayes factor: log(BF_10) = log p(Y|M_1) - log p(Y|M_0)
  log_bf <- log_ml_nonstationary - log_ml_stationary
  
  # Bayes factor on original scale
  bf <- exp(log_bf)
  
  # Interpretation based on Kass & Raftery (1995) guidelines
  interpretation <- case_when(
    log_bf > 4.6 ~ "Decisive for nonstationary",
    log_bf > 2.3 ~ "Strong for nonstationary",
    log_bf > 1.0 ~ "Substantial for nonstationary",
    log_bf > 0 ~ "Weak for nonstationary",
    log_bf > -1.0 ~ "Weak for stationary",
    log_bf > -2.3 ~ "Substantial for stationary",
    log_bf > -4.6 ~ "Strong for stationary",
    TRUE ~ "Decisive for stationary"
  )
  
  return(list(
    log_bayes_factor = log_bf,
    bayes_factor = bf,
    interpretation = interpretation
  ))
}
```

### Main Analysis Function

```{r main-analysis-function}
# Function to analyze a single gene
analyze_gene_nonstationarity <- function(coords, expr, gene_name = NULL, 
                                         normalize = TRUE, detrend = TRUE,
                                         detrend_method = "loess",
                                         verbose = FALSE) {
  
  # Check INLA availability
  if (!exists(".inla_available") || !get(".inla_available")) {
    if (!require("INLA", quietly = TRUE)) {
      stop("INLA package is required but not installed. Please install it using:\n",
           "install.packages('INLA', repos=c(getOption('repos'), INLA='https://inla.r-inla-download.org/R/stable'), dep=TRUE)\n",
           "Then RESTART R/RStudio and try again.")
    }
    # Test if binaries are loaded
    tryCatch({
      INLA::inla.mesh.1d(seq(0, 1, length.out = 5))
    }, error = function(e) {
      stop("INLA is installed but binaries are not loading. Error: ", e$message, "\n",
           "SOLUTION: Please RESTART R/RStudio completely. INLA requires a restart after installation.")
    })
  }
  
  # Step 1: Normalize expression
  if (normalize) {
    expr_processed <- normalize_expression(expr, method = "quantile")
  } else {
    expr_processed <- expr
  }
  
  # Step 2: Detrend expression
  if (detrend) {
    detrend_result <- detrend_expression(coords, expr_processed, method = detrend_method)
    expr_processed <- detrend_result$detrended
  }
  
  # Create mesh (shared between models)
  mesh <- create_spde_mesh(coords)
  
  # Step 3: Fit stationary model
  if (verbose) cat("Fitting stationary model...\n")
  fit_stationary <- fit_stationary_matern(coords, expr_processed, mesh = mesh, verbose = verbose)
  
  # Step 4: Fit nonstationary model
  if (verbose) cat("Fitting nonstationary model...\n")
  fit_nonstationary <- fit_nonstationary_matern_simple(coords, expr_processed, mesh = mesh, verbose = verbose)
  
  # Step 5: Extract results
  results_stationary <- extract_variance_components(fit_stationary$result)
  results_nonstationary <- extract_variance_components(fit_nonstationary$result)
  
  # Step 6: Calculate Bayes factor
  bf_result <- calculate_bayes_factor(
    results_stationary$log_marginal_likelihood,
    results_nonstationary$log_marginal_likelihood
  )
  
  # Compile results
  results <- data.frame(
    gene_name = ifelse(is.null(gene_name), "unknown", gene_name),
    # Stationary model results
    sigma_b_sq_stationary = results_stationary$sigma_b_sq,
    sigma_eps_sq_stationary = results_stationary$sigma_eps_sq,
    range_stationary = results_stationary$spatial_range,
    prop_spatial_stationary = results_stationary$prop_spatial,
    log_ml_stationary = results_stationary$log_marginal_likelihood,
    # Nonstationary model results
    sigma_b_sq_nonstationary = results_nonstationary$sigma_b_sq,
    sigma_eps_sq_nonstationary = results_nonstationary$sigma_eps_sq,
    range_nonstationary = results_nonstationary$spatial_range,
    prop_spatial_nonstationary = results_nonstationary$prop_spatial,
    log_ml_nonstationary = results_nonstationary$log_marginal_likelihood,
    # Bayes factor: log(BF_10) = log p(Y|M_1) - log p(Y|M_0)
    log_bayes_factor = bf_result$log_bayes_factor,
    bayes_factor = bf_result$bayes_factor,
    bf_interpretation = bf_result$interpretation,
    stringsAsFactors = FALSE
  )
  
  return(list(
    results = results,
    fits = list(
      stationary = fit_stationary,
      nonstationary = fit_nonstationary
    )
  ))
}

# Function to analyze multiple genes
analyze_multiple_genes <- function(spe, gene_indices, normalize = TRUE, 
                                   detrend = TRUE, detrend_method = "loess",
                                   verbose = FALSE) {
  
  # Check INLA availability
  if (!exists(".inla_available") || !get(".inla_available")) {
    if (!require("INLA", quietly = TRUE)) {
      stop("INLA package is required but not installed. Please install it using:\n",
           "install.packages('INLA', repos=c(getOption('repos'), INLA='https://inla.r-inla-download.org/R/stable'), dep=TRUE)\n",
           "Then RESTART R/RStudio and try again.")
    }
    # Test if binaries are loaded
    tryCatch({
      INLA::inla.mesh.1d(seq(0, 1, length.out = 5))
    }, error = function(e) {
      stop("INLA is installed but binaries are not loading. Error: ", e$message, "\n",
           "SOLUTION: Please RESTART R/RStudio completely. INLA requires a restart after installation.")
    })
  }
  
  # Get coordinates
  coords <- as.data.frame(spatialCoords(spe))
  colnames(coords)[1:2] <- c("x", "y")
  
  # Get gene names
  rd <- rowData(spe)
  gene_names <- if ("symbol" %in% colnames(rd)) {
    rd$symbol[gene_indices]
  } else if ("gene_name" %in% colnames(rd)) {
    rd$gene_name[gene_indices]
  } else {
    rownames(spe)[gene_indices]
  }
  
  # Initialize results list
  all_results <- list()
  results_df <- data.frame()
  
  # Process each gene
  for (i in seq_along(gene_indices)) {
    gene_idx <- gene_indices[i]
    gene_name <- gene_names[i]
    
    if (verbose) {
      cat(sprintf("Processing gene %d/%d: %s\n", i, length(gene_indices), gene_name))
    }
    
    # Get expression
    expr <- log1p(counts(spe)[gene_idx, ])
    
    # Analyze gene
    tryCatch({
      analysis_result <- analyze_gene_nonstationarity(
        coords = coords,
        expr = expr,
        gene_name = gene_name,
        normalize = normalize,
        detrend = detrend,
        detrend_method = detrend_method,
        verbose = verbose
      )
      
      # Store results
      all_results[[gene_name]] <- analysis_result
      results_df <- rbind(results_df, analysis_result$results)
      
    }, error = function(e) {
      warning(sprintf("Error analyzing gene %s: %s\n", gene_name, e$message))
      # Add NA row for failed gene (must match columns in results dataframe)
      na_row <- data.frame(
        gene_name = gene_name,
        sigma_b_sq_stationary = NA_real_,
        sigma_eps_sq_stationary = NA_real_,
        range_stationary = NA_real_,
        prop_spatial_stationary = NA_real_,
        log_ml_stationary = NA_real_,
        sigma_b_sq_nonstationary = NA_real_,
        sigma_eps_sq_nonstationary = NA_real_,
        range_nonstationary = NA_real_,
        prop_spatial_nonstationary = NA_real_,
        log_ml_nonstationary = NA_real_,
        log_bayes_factor = NA_real_,
        bayes_factor = NA_real_,
        bf_interpretation = NA_character_,
        stringsAsFactors = FALSE
      )
      results_df <<- rbind(results_df, na_row)
    })
  }
  
  # Reset row names to be sequential
  if (nrow(results_df) > 0) {
    rownames(results_df) <- 1:nrow(results_df)
  }
  
  return(list(
    results_df = results_df,
    detailed_results = all_results
  ))
}
```

## Example Analysis

```{r example-analysis}
# Load example dataset
# Query ExperimentHub for the dataset
visium_hbc <- query(eh, "HumanBreastCancerIDC")

# Check what we got
if (length(visium_hbc) == 0) {
  stop("No datasets found matching 'HumanBreastCancerIDC'. Please check the dataset name.")
}

# Extract the SpatialExperiment object
# The [[1]] accesses the first (and typically only) resource
spe <- visium_hbc[[1]]

# Verify it's a SpatialExperiment object
if (!inherits(spe, "SpatialExperiment")) {
  stop("Expected SpatialExperiment object, but got: ", class(spe))
}

# Select genes with actual expression (not the first genes which are often non-expressed)
# Calculate total expression per gene and select top expressed genes
gene_totals <- Matrix::rowSums(counts(spe))
# Get indices of genes with highest expression
top_genes <- order(gene_totals, decreasing = TRUE)

# Select top 10 expressed genes for testing
n_genes <- 10
gene_indices <- top_genes[1:n_genes]

# Show which genes we're analyzing
rd <- rowData(spe)
gene_names <- if ("symbol" %in% colnames(rd)) {
  rd$symbol[gene_indices]
} else if ("gene_name" %in% colnames(rd)) {
  rd$gene_name[gene_indices]
} else {
  rownames(spe)[gene_indices]
}
cat("Analyzing top expressed genes:\n")
print(data.frame(index = gene_indices, name = gene_names, total_counts = gene_totals[gene_indices]))

# Run analysis
analysis_results <- analyze_multiple_genes(
  spe = spe,
  gene_indices = gene_indices,
  normalize = TRUE,
  detrend = TRUE,
  detrend_method = "loess",
  verbose = TRUE
)

# Display results
print(analysis_results$results_df)
```

```{r}
analysis_results$results_df[,c("gene_name","log_bayes_factor","bf_interpretation")]
```

## Results Summary

```{r results-summary}
# Summary statistics
summary_stats <- analysis_results$results_df %>%
  summarise(
    n_genes = n(),
    n_successful = sum(!is.na(bayes_factor)),
    prop_favor_nonstationary = mean(favors_nonstationary, na.rm = TRUE),
    mean_log_bf = mean(log_bayes_factor, na.rm = TRUE),
    median_log_bf = median(log_bayes_factor, na.rm = TRUE),
    mean_prop_spatial_stationary = mean(prop_spatial_stationary, na.rm = TRUE),
    mean_prop_spatial_nonstationary = mean(prop_spatial_nonstationary, na.rm = TRUE),
    mean_diff_prop_spatial = mean(prop_spatial_nonstationary - prop_spatial_stationary, na.rm = TRUE)
  )

print(summary_stats)

# Distribution of Bayes factors
ggplot(analysis_results$results_df, aes(x = log_bayes_factor)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Log Bayes Factors",
    subtitle = "Positive values favor nonstationary model",
    x = "Log Bayes Factor",
    y = "Frequency"
  ) +
  theme_minimal()

# Comparison of spatial variability proportions
comparison_df <- analysis_results$results_df %>%
  select(gene_name, prop_spatial_stationary, prop_spatial_nonstationary) %>%
  pivot_longer(
    cols = c(prop_spatial_stationary, prop_spatial_nonstationary),
    names_to = "model_type",
    values_to = "prop_spatial"
  ) %>%
  mutate(model_type = ifelse(model_type == "prop_spatial_stationary", 
                             "Stationary", "Nonstationary"))

ggplot(comparison_df, aes(x = model_type, y = prop_spatial, fill = model_type)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Stationary" = "steelblue", "Nonstationary" = "coral")) +
  labs(
    title = "Comparison of Spatial Variability Proportions",
    subtitle = "Stationary vs Nonstationary Models",
    x = "Model Type",
    y = "Proportion of Spatial Variability"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Notes on Implementation

The code above implements the analysis framework described. Key points:

1.  **Normalization**: Quantile normalization is used by default to make gene expression comparable across genes.

2.  **Detrending**: Loess smoothing is used to remove large-scale spatial trends before modeling the residual spatial structure.

3.  **Stationary Model**: Uses a standard Matern SPDE model with constant covariance parameters.

4.  **Nonstationary Model**: Uses a simplified approach where spatial covariates interact with the spatial field to capture nonstationarity. For a fully nonstationary Matern model with spatially varying range and variance parameters, more advanced INLA techniques would be needed (see INLA documentation on nonstationary SPDE models).

5.  **Bayes Factor**: Calculated from the marginal likelihoods provided by INLA to compare stationary vs nonstationary models.

6.  **Spatial Variability Proportion**: Calculated as σ_b²/(σ_b² + σ_ε²) for both models.

The results can be summarized across genes, datasets, tissue types, and gene groups as needed for the research questions.
